Why Do People Punish? Evidence for a Range of Strategic Concerns

### 2025-08-01 quarto/manuscript.qmd
Update manuscript and submit

---
title             : "Why do people punish? Evidence for a range of strategic concerns"
shorttitle        : "Punishment strategies"

author: 
  - name          : "Scott Claessens"
    affiliation   : "*1"
  - name          : "Quentin D. Atkinson"
    affiliation   : "1"
  - name          : "Nichola Raihani"
    affiliation   : "1,2"

affiliation:
  - id            : "1"
    institution   : "\\footnotesize School of Psychology, University of Auckland, Auckland, New Zealand"
  - id            : "2"
    institution   : "\\footnotesize Department of Experimental Psychology, University College London, London, United Kingdom"

abstract: |
 Costly punishment is thought to be a key mechanism sustaining human cooperation. However, the motives for punitive behaviour remain unclear. Although often assumed to be motivated by a desire to convert cheats into cooperators, punishment is also consistent with other functions, such as levelling payoffs or improving one’s relative position. We used six economic games to tease apart different motives for punishment and to explore whether different punishment strategies were associated with personality variables, political ideology, and religiosity. We used representative samples from the United Kingdom and the United States (N = 2010) to estimate the frequency of different punishment strategies in the population. The most common strategy was to never punish: prosocial individuals and those scoring high in honesty-humility were among the least punitive in our sample, while religious individuals and those scoring high for right-wing authoritarianism and social dominance orientation were more punitive. For people who did punish, strategy use was more consistent with egalitarian motives than behaviour-change motives. Nevertheless, different punishment strategies were also associated with personality, social preferences, political ideology, and religiosity. Self-reports of behaviour in the games suggested that people have some insight into their punishment strategy. These findings highlight the multipurpose nature of human punishment and show how the different motives underpinning punishment decisions are linked with core character traits.

bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{makecell}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This manuscript has not yet been peer-reviewed. \par This manuscript was published as a pre-print on PsyArXiv under a CC BY 4.0 \text{license:} \url{https://doi.org/10.31234/osf.io/ys6rm} \par \text{Classification:} Social Sciences; Psychological and Cognitive Sciences \par \text{Keywords:} punishment; cooperation; economic games}
appendix          : "appendix.Rmd"
csl               : pnas.csl
---

```{r options, echo=FALSE}
opts_chunk$set(echo = FALSE)
```

\linenumbers

# Introduction

Humans cooperate on a scale that is unparalleled in the animal kingdom. One 
mechanism thought to sustain this level of cooperation is costly punishment, 
whereby individuals harm others at a personal cost [@CluttonBrock1995], 
ostensibly encouraging cooperative behaviour from the target, or bystanders 
[@DosSantos2011;@DosSantos2013;@Raihani2015], in the future. Punishment offers a 
route to maintaining or increasing cooperation by changing the payoff structure 
of social interactions such that it no longer pays to cheat or exploit social 
partners [@CluttonBrock1995;@Ostrom1990]. Yet, despite its theoretical 
importance, the question of why people choose to punish others is still hotly 
contested [@Raihani2019]. In this study, we use a battery of economic games to 
disentangle the different motives underpinning punishment and explore how these 
motives vary across individuals.

In humans, many studies of punishment have been carried out in laboratory 
settings using economic games [@Balliet2011;@Balliet2013;@Chaudhuri2011;
@Dreber2008;@Fehr2000;@Fehr2002;@Henrich2006;@Nikiforakis2008a;@Raihani2012a;
@Raihani2019]. In these games, participants are usually given a sum of money 
that they can use to invest in collective action or to help others. 
Alternatively, participants can 'cheat' by keeping the money for themselves or 
by exploiting the contributions of others. Punishment is introduced into such 
games by giving participants the option to pay a small 'fee' to impose a greater 
'fine' on their co-players. Several lines of experimental evidence indicate that 
people use this punishment option [@Henrich2006], that they enjoy 
punishing [@DeQuervain2004], and that they frequently, though not 
always [@Herrmann2008], punish cheating or exploitative 
co-players [@Fehr2000;@Fehr2002].

Evidence from these experiments suggests that the threat of costly punishment 
plays an important role in promoting human cooperation. People tend to cooperate 
more in games where punishment is possible compared to those where it is not
[@Balliet2011;@Balliet2013;@Raihani2019]. The effect that the threat of 
punishment has on cooperation is also evident in the higher contributions 
typically observed in the Ultimatum Game (where punishment is possible) compared 
to the structurally-similar Dictator Game (where it is not) [@Camerer2003]. This 
typical cooperation-enhancing effect of punishment has also been observed across 
societies [@Balliet2013], leading some to suggest that costly punishment has 
played a key role in the cultural evolution of cooperation in humans
[@Bowles2013;@Boyd2003;@Chudek2011;@Henrich2017].

Nevertheless, it remains unclear whether individuals playing economic games use
punishment as a behaviour-change tool to enforce cooperation or as a means to 
achieve other ends. Some have argued that punishment is primarily used to shape 
future behaviour, either to deter personal harm [@Delton2017;@Dreber2008;
@DosSantos2013] or to uphold normative standards of cooperative behaviour 
[@Boyd2003;@Chudek2011;@Fehr2018;@Mathew2011;@Mathew2014;@Richerson2016]. 
But while the _threat_ of punishment can have a cooperation-enhancing effect, 
the _enactment_ of this punishment does not consistently deter targets from 
cheating in the future [@Raihani2019]. This calls into question whether 
punishment primarily operates as a behaviour-change tool or whether it is used 
to achieve other goals.

Beyond behaviour shaping concerns, there are a host of other reasons why people 
may want to punish in economic games. Punishers might be motivated by a desire 
for retribution rather than deterrence, punishing in proportion to the amount 
of harm that was personally caused [@Carlsmith2002]. Punishment might be driven 
by concerns about relative payoffs, such as disadvantageous inequity aversion 
(i.e., avoiding having less than others) [@Raihani2012b;@Raihani2019] and/or 
general egalitarian preferences (i.e., wanting all participants to receive the 
same payoffs) [@Dawes2007]. Such concerns about relative payoffs may be activated 
when participants earn less than cheats in economic games or when there are 
income disparities in these settings. People might also use punishment for 
competitive purposes, seeking advantageous inequity for themselves (i.e., having 
more than others) and/or improving their relative position [@Raihani2019].

Common economic game designs have been unable to tease apart these different 
motives for punishment because participants who interact with cheats in these 
games experience both losses _and_ lower relative payoffs. The typical 1:3 
fee-fine ratio of punishment in economic games compounds this issue. With 
this setup, people can simultaneously use punishment to reciprocate losses, to 
deter others from cheating, and to reduce or reverse disparities in payoffs 
between themselves and targets. To add to this complexity, it is evident that 
people use punishment in seemingly disparate ways: punishing when no behaviour 
change is possible, such as in one-shot games [@Bone2015;@Henrich2006;
@Raihani2012b;@Walker2004], on the very last round of repeated games
[@Barclay2016], or in games where the target never learns about the 
punishment [@Crockett2014]; punishing those who did not cheat or who 
over-contributed to collective action (antisocial punishment) 
[@Herrmann2008;@Sylwester2013]; punishing in scenarios where they were not 
personally harmed (third-party punishment) [@Fehr2004]; and punishing in 
scenarios where disparities in payoffs did not arise from participants' 
actions [@Dawes2007;@Johnson2009;@Fowler2005].

@@

This insight raises several underexplored questions. First, which punishment 
strategies are more frequent in human populations? Second, what traits predict 
adherence to a particular strategy? Previous work has reported that personality 
is related to cooperative behaviour [@Thielmann2020] and demographics, political 
ideology, and religiosity are related to punitive behaviour [@Hofmann2018], but 
no research has related these variables to specific punishment strategies. 
Third, do people have insight into their own punishment strategy? Previous work 
has argued that people are often unaware of the underlying function of their 
punitive behaviour, yet they feel compelled to enact it anyway
[@Carlsmith2002;@Nisbett1977].

Here, we aim to delineate nine possible punishment strategies by asking 
whether people punish in a manner consistent with a specific strategy and, if 
so, what other characteristics (personality, social preferences, political 
orientation, religious views) predict the use of different punishment strategies.
Table \@ref(tab:tableStrategies) summarises the potential functions for costly 
punishment in the economic games that we considered, and the behavioural 
strategies they predict. Note that Table \@ref(tab:tableStrategies) is not 
an exhaustive list of all possible punishment strategies: we do not include 
reputational functions of punishment in this table, such as signalling 
trustworthiness [@Barclay2006;@Batistoni2022;@Jordan2016;@Jordan2017;@Raihani2015], 
because our focus is on punishment strategies in anonymous economic games without 
reputational incentives (but see ref[@Jordan2020]).

Building on previous designs [@Bone2015;@Deutchman2021;@Marczyk2017;
@Raihani2012b], we employ a suite of one-shot economic games where individuals 
are given the opportunity to punish targets at a personal cost (Figure 
\@ref(fig:plotGames)). In each game, targets either steal from another 
individual or do nothing. Representative samples of participants from the United 
Kingdom (*n* = `r sum(tar_read(dExc)$Country == "United Kingdom")`) and the 
United States (*n* = `r sum(tar_read(dExc)$Country == "United States")`) 
completed all six games on the online platform Prolific. We carefully designed 
the suite of games to tease apart the proposed punishment strategies in Table 
\@ref(tab:tableStrategies), such that each strategy predicts a different pattern 
of behaviour across all the games (see Methods for more detail about the six 
games). We use the resulting behavioural patterns to discern which punishment 
strategy participants are employing. We then combine these behavioural patterns 
with data on demographics, personality, social preferences, political ideology,
religiosity, and self-reported strategy usage.

(ref:plotGamesCaption) *Visual summary of the six economic games.* In the
games, Player 2 either steals £0.20 from Player 1 (the focal player) or does 
nothing. Player 1 is then given the option to punish by paying a certain amount 
of money to remove money from Player 2 (this money is destroyed). The six games
are variants on this general setup, creating situations where (A) Player 2 is
still worse off after stealing, (B) Player 2 creates equality by stealing, (C) 
the computer "decides" whether Player 2 steals, (D) the fee-fine ratio is 1:1, 
(E) Player 2 is better off after stealing, and (F) Player 2 steals instead from 
a third-party.

```{r plotGames, out.width="100%", fig.cap="(ref:plotGamesCaption)"}
include_graphics("images/games.png")
```

(ref:tableStrategiesCaption) \scriptsize Summary of the different functions for 
punishment and the behavioural strategies they predict. *Games A-F are 
the games employed in the current study (see Methods for more details). In each 
of the six games, participants are given the opportunity to punish players who 
"steal" and those who do not, meaning that participants make twelve punishment 
decisions in total. Each behavioural strategy implies a unique pattern of 
punishment across all decisions (see Methods for detailed explanation of 
strategies). Green ticks reflect decisions to punish, red crosses reflect 
decisions to not punish. In column headers, payoffs at the first stage (above) 
and the second stage (below) are denoted as P1-P2 (or P2-P3 [P1] for Game F) 
where participants take the role of P1 and P2 is the target of punishment. AI = 
advantageous inequity, DI = disadvantageous inequity.*

```{r tableStrategies}
apa_table(
  tar_read(tableStrategies),
  escape = FALSE,
  landscape = TRUE,
  align = c("p{1.7cm}", "p{3.3cm}", rep("c", 12)),
  col_spanners = list(
    "\\thead{\\scriptsize Game A \\\\ (AI) \\\\ 70-10}"   = 3:4,
    "\\thead{\\scriptsize Game B \\\\ (Equal) \\\\ 70-30}"   = 5:6,
    "\\thead{\\scriptsize Game C \\\\ (Computer) \\\\ 70-30}"   = 7:8,
    "\\thead{\\scriptsize Game D \\\\ (1:1 Fee-Fine) \\\\ 70-30}"   = 9:10,
    "\\thead{\\scriptsize Game E \\\\ (DI) \\\\ 70-50}"        = 11:12,
    "\\thead{\\scriptsize Game F \\\\ (Third-Party) \\\\ 70-70 [100]}" = 13:14
    ),
  caption = "(ref:tableStrategiesCaption)",
  font_size = "scriptsize"
)
```

@@

The overall pattern of punitive behaviour in the six economic games was very 
similar across both countries (Figure \@ref(fig:plotDecisions)). Participants 
were generally more likely to punish targets who stole compared to targets who 
did not steal (multilevel logistic regression; _b_ = 
`r printnum(mlm1["otherTook","Estimate"])`, standard error = 
`r printnum(mlm1["otherTook","Std. Error"])`, _p_ 
`r printp(mlm1["otherTook","Pr(>|z|)"])`). Participants were also more likely 
to punish when targets' stealing behaviour generated inequalities, 
specifically in Games E and F (_b_ = 
`r printnum(mlm2["DI","Estimate"])`, SE = `r printnum(mlm2["DI","Std. Error"])`, 
_p_ `r printp(mlm2["DI","Pr(>|z|)"])`).

(ref:plotDecisionsCaption) *Overall pattern of punitive behaviour across all six 
economic games, split by country.* AI = advantageous inequity, DI = 
disadvantageous inequity.

```{r plotDecisions, fig.width=8, fig.height=5, fig.cap="(ref:plotDecisionsCaption)"}
tar_read(plotDecisions)
```

We classified participants into a particular strategy if their behaviour across 
all twelve decisions matched our behavioural predictions shown in Table 
\@ref(tab:tableStrategies) exactly. Table \@ref(tab:tableStrategyCounts) shows 
the proportion of participants following each strategy, with N/A used to 
represent participants who did not fit exactly into any particular strategy type.
Overall, `r printnum(mean(tar_read(dExc)$strategy != "N/A")*100, digits = 0)`%
of our participants could be classified exactly into one of the strategies. The
most common strategy in both countries was to never punish across any of the
games. The next most common strategies were those that care about
minimising payoff differences (avoid disadvantageous inequity, egalitarian). 
Less common were the behaviour shaping strategies (deterrent, norm-enforcing),
the retributive strategy, and the competitive strategies (seek advantageous 
inequity, competitive). Although participants often punished targets who did not 
steal in the six games (Figure \@ref(fig:plotDecisions)), no participants 
followed the antisocial strategy by exclusively punishing targets who did not 
steal across _all_ games.

(ref:tableStrategyCountsCaption) Counts and proportions of participants 
following each punishment strategy exactly, split by country. *N/A implies that
participants were unable to be classified exactly into any of the punishment
strategies. AI = advantageous inequity, DI = disadvantageous inequity.*

```{r tableStrategyCounts}
apa_table(
  tar_read(tableStrategyCounts),
  escape = FALSE,
  align = c("l", rep("c", 4)),
  col_spanners = list(
    "\\thead{United Kingdom \\\\ (N = 1014)}" = 2:3,
    "\\thead{United States \\\\ (N = 996)}" = 4:5
    ),
  caption = "(ref:tableStrategyCountsCaption)"
)
```

To further investigate the strategies that participants were following, we 
examined the most common patterns of punitive behaviour across all twelve 
decisions. Supplementary Table \@ref(tab:tablePatterns) shows the proportion of 
participants following the 25 most common behavioural patterns, including, 
where appropriate, the predetermined strategies from Table 
\@ref(tab:tableStrategies). In both countries, a common pattern of behaviour 
not captured by any of the strategies was punishing only when the target stole
in the third-party game (Game F). Punishment in this game is consistent with an 
egalitarian motive, as stealing produces unequal outcomes, but third-party 
punishment here is also consistent with norm-enforcing and competitive motives 
(see Table \@ref(tab:tableStrategies)). Other common behavioural patterns not 
captured by our strategies included punishing whenever the target stole across 
all games and always punishing in every game irrespective of the targets' 
behaviour.

@@

To deal with this complexity and include all observed data in our frequency 
estimates, we fitted a Bayesian latent state model to the data. This model 
assumes that the nine strategies in Table \@ref(tab:tableStrategies) (plus a 
"random choice" strategy that chooses randomly for each decision) are the only 
latent strategies and that these are instantiated into observed behaviour 
according to the logic in Table \@ref(tab:tableStrategies) with some 
probability of implementation error (i.e., an intention to punish is implemented 
as non-punishment and vice versa). Averaging over all strategies and 
incorporating the possibility of implementation errors, the model estimates the 
probability of participants following any particular strategy, given the 
observed data.

The posterior estimates from the model are presented in Figure 
\@ref(fig:plotModel1b). The posterior probabilities for each strategy did not
differ between the two countries. In both countries, the never punish strategy
had the highest probability, followed by the egalitarian strategy. The
norm-enforcing and seek advantageous inequity strategies were the next most 
likely, with higher posterior estimates than the competitive and antisocial 
strategies. None of the other strategies differed in their posterior estimates. 
The same general pattern emerged when we analysed the full dataset without 
pre-registered exclusions (Supplementary Figure \@ref(fig:plotModel1a)).

>>>>

---
title: "Why Do People Punish? Evidence for a Range of Strategic Concerns"
shorttitle: "Punishment strategies"
author:
  - name: Scott Claessens
    corresponding: true
    orcid: 0000-0002-3562-6981
    email: scott.claessens@gmail.com
    affiliations:
      - name: University of Kent
        department: School of Psychology
        city: Canterbury
        country: United Kingdom
      - name: University of Auckland
        department: School of Psychology
        city: Auckland
        country: New Zealand
    role:
      - conceptualization
      - data curation
      - formal analysis
      - investigation
      - methodology
      - visualization
      - writing
      - editing
  - name: Quentin D Atkinson
    corresponding: false
    orcid: 0000-0002-8499-7535
    affiliations:
      - name: University of Auckland
        department: School of Psychology
        city: Auckland
        country: New Zealand
    role:
      - conceptualization
      - funding acquisition
      - methodology
      - supervision
      - writing
      - editing
  - name: Nichola J Raihani
    corresponding: false
    orcid: 0000-0003-2339-9889
    affiliations:
      - name: University of Auckland
        department: School of Psychology
        city: Auckland
        country: New Zealand
      - name: University College London
        department: Department of Experimental Psychology
        city: London
        country: United Kingdom
    role:
      - conceptualization
      - funding acquisition
      - investigation
      - methodology
      - supervision
      - writing
      - editing
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    study-registration: "This manuscript was published as a pre-print on PsyArXiv under a CC BY 4.0 license: <https://doi.org/10.31234/osf.io/ys6rm>"
    data-sharing: Word count (excluding abstract, figures, tables, and references) = 6065 words.
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
abstract: "Costly punishment is thought to be a key mechanism sustaining human cooperation. However, the motives for punitive behaviour remain unclear. Although often assumed to be motivated by a desire to convert cheats into cooperators, punishment is also consistent with other functions, such as levelling payoffs or improving one’s relative position. We used six economic games to tease apart different motives for punishment and to explore whether different punishment strategies were associated with personality variables, political ideology, and religiosity. We used representative samples from the United Kingdom and the United States (N = 2010) to estimate the frequency of different punishment strategies in the population. The most common strategy was to never punish: prosocial individuals and those scoring high in honesty-humility were among the least punitive in our sample, while religious individuals and those scoring high for right-wing authoritarianism and social dominance orientation were more punitive. For people who did punish, strategy use was more consistent with egalitarian motives than behaviour-change motives. Nevertheless, different punishment strategies were also associated with personality, social preferences, political ideology, and religiosity. Self-reports of behaviour in the games suggested that people have some insight into their punishment strategy. These findings highlight the multipurpose nature of human punishment and show how the different motives underpinning punishment decisions are linked with core character traits."
keywords: [punishment, cooperation, economic games]
bibliography: bibliography.bib
engine: knitr
format:
  apaquarto-docx:
    fig-dpi: 300
execute:
  echo: false
  warning: false
  error: false
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Supplementary Figure
      space-before-numbering: true
      latex-list-of-description: Supplementary Figure
      caption-location: bottom
    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Supplementary Table
      space-before-numbering: true
      latex-list-of-description: Supplementary Table
      caption-location: top
floatsintext: true
csl: nature.csl
---

```{r}
library(flextable)
library(knitr)
library(targets)
library(tidyverse)
```

Humans cooperate on a scale that is unparalleled in the animal kingdom. One 
mechanism thought to sustain this level of cooperation is costly punishment, 
whereby individuals harm others at a personal cost [@CluttonBrock1995], 
ostensibly encouraging cooperative behaviour from the target or bystanders 
[@DosSantos2011; @DosSantos2013; @Raihani2015]. Punishment offers a route to 
maintaining or increasing cooperation by changing the payoff structure of social 
interactions such that it no longer pays to cheat or exploit social partners 
[@CluttonBrock1995; @Ostrom1990]. Yet, despite its theoretical importance, the 
question of why people choose to punish others is still hotly contested 
[@Raihani2019]. In this study, we use a battery of economic games to disentangle
the different motives underpinning punishment and explore how these motives vary 
across individuals.

In humans, punishment is often studied in laboratory settings using economic 
games [@Balliet2011; @Balliet2013; @Chaudhuri2011; @Dreber2008; @Fehr2000; 
@Fehr2002; @Henrich2006; @Nikiforakis2008a; @Raihani2012a; @Raihani2019]. In 
some of these games, participants are given a sum of money that they can use to 
invest in collective action or to help others; conversely, they can 'cheat' by 
keeping the money for themselves or by exploiting the contributions of others. 
In other games, participants are endowed with sums of money and can then 'cheat'
by destroying a partner's payoffs or stealing from them [@Abbink2009; 
@Raihani2012b]. Punishment is introduced into such games by giving participants 
the option to pay a small 'fee' to impose a greater 'fine' on their co-players. 
Several lines of experimental evidence indicate that people use this punishment 
option [@Henrich2006], that they enjoy punishing [@DeQuervain2004], and that 
they frequently, though not always [@Herrmann2008], punish those who cheat,
free-ride, or steal [@Fehr2000; @Fehr2002].

Evidence from these experiments suggests that the threat of costly punishment 
plays an important role in promoting human cooperation. People tend to cooperate 
more in games where punishment is possible compared to those where it is not
[@Balliet2011; @Balliet2013; @Raihani2019]. Similarly, people make higher offers 
in the Ultimatum Game (where punishment is possible) compared to the 
structurally-similar Dictator Game (where it is not) [@Camerer2003]. This 
cooperation-enhancing effect of punishment has also been observed across 
societies [@Balliet2013], leading some to suggest that costly punishment has 
played a key role in the cultural evolution of cooperation in humans
[@Bowles2013; @Boyd2003; @Chudek2011; @Henrich2017].

Nevertheless, it remains unclear whether individuals playing economic games use
punishment as a behaviour-change tool to enforce cooperation or to achieve other 
ends. Some have argued that punishment is primarily used to shape behaviour, 
either to deter personal harm [@Delton2017; @Dreber2008; @DosSantos2013] or to 
uphold normative standards of cooperative behaviour [@Boyd2003; @Chudek2011; 
@Fehr2018; @Mathew2011; @Mathew2014; @Richerson2016]. But while the _threat_ of 
punishment can have a cooperation-enhancing effect, the _enactment_ of this 
punishment does not consistently deter targets from cheating in the future 
[@Raihani2019]. This calls into question whether punishment primarily operates 
as a behaviour-change tool or whether it is used to achieve other goals.

Beyond behaviour shaping concerns, there are other reasons why people punish in 
economic games. Punishers might be motivated by a desire for retribution rather 
than deterrence, punishing in proportion to the amount of harm that was 
personally caused [@Carlsmith2002]. Punishment might be driven by concerns about 
relative payoffs, such as disadvantageous inequity aversion (i.e., avoiding 
having less than others) [@Raihani2012b; @Raihani2019] and/or general 
egalitarian preferences (i.e., wanting all participants to receive the same 
payoffs) [@Dawes2007]. Such concerns about relative payoffs may be activated 
when participants earn less than cheats in economic games or when there are 
income disparities in these settings. People might also use punishment for 
competitive purposes, seeking advantageous inequity for themselves (i.e., having 
more than others) and/or improving their relative position [@Raihani2019].

Common economic game designs have been unable to tease apart these different 
motives for punishment because participants who interact with cheats in these 
games experience both losses _and_ lower relative payoffs. The typical 1:3 
fee-fine ratio of punishment in economic games compounds this issue. With 
this setup, people can simultaneously use punishment to reciprocate losses, to 
deter others from cheating, and to reduce or reverse disparities in payoffs 
between themselves and targets. To add to this complexity, it is evident that 
people use punishment in seemingly disparate ways: punishing when no behaviour 
change is possible, such as in one-shot games [@Bone2015; @Henrich2006;
@Raihani2012b; @Walker2004], on the very last round of repeated games 
[@Barclay2016], or in games where the target never learns about the punishment 
[@Crockett2014]; punishing those who did not cheat or who over-contributed to 
collective action (sometimes called 'antisocial punishment') [@Herrmann2008; 
@Sylwester2013]; punishing in scenarios where they were not personally harmed 
(third-party punishment) [@Fehr2004]; and punishing in scenarios where 
disparities in payoffs did not arise from participants' actions [@Dawes2007; 
@Johnson2009; @Fowler2005].

@@

This insight raises several underexplored questions. First, which punishment 
strategies are more frequent in human populations? Second, what traits predict 
adherence to a particular strategy? Previous work has reported that personality 
is related to cooperative behaviour [@Thielmann2020] and demographics, political 
ideology, and religiosity are related to punitive behaviour [@Hofmann2018], but 
no research has related these variables to specific punishment strategies. 
Third, do people have insight into their own punishment strategy? Previous work 
has argued that people are often unaware of the underlying function of their 
punitive behaviour, yet they feel compelled to enact it anyway
[@Carlsmith2002; @Nisbett1977].

Here, we aim to delineate nine possible punishment strategies by asking 
whether people punish in a manner consistent with a specific strategy and, if 
so, what other characteristics (personality, social preferences, political 
orientation, religious views) predict the use of different punishment 
strategies. @tbl-strategies summarises the potential functions for costly 
punishment in the economic games that we considered, and the behavioural 
strategies they predict. Note that @tbl-strategies is not an exhaustive list of 
all possible punishment strategies: we do not include reputational functions of 
punishment in this table, such as signalling trustworthiness [@Barclay2006; 
@Batistoni2022; @Jordan2016; @Jordan2017; @Raihani2015], because our focus is on 
punishment strategies in anonymous economic games without reputational 
incentives (but see ref [@Jordan2020]).

Building on previous designs that have used one-shot economic games to explore 
behaviour-shaping and egalitarian motives for punishment [@Bone2015; 
@Crockett2014; @Deutchman2021; @Marczyk2017; @Raihani2012b], we employ a suite 
of one-shot economic games where individuals are given the opportunity to punish 
targets at a personal cost (@fig-games). In each game, targets either steal from 
another individual or do nothing. Representative samples of participants from 
the United Kingdom (*n* = `r sum(tar_read(dExc)$Country == "United Kingdom")`) 
and the United States (*n* = `r sum(tar_read(dExc)$Country == "United States")`) 
completed all six games on the online platform Prolific. We designed the suite 
of games to tease apart the proposed punishment strategies in @tbl-strategies, 
such that each strategy predicts a different pattern of behaviour across all the 
games (see Methods for more detail about the six games). We use the resulting 
behavioural patterns to discern which punishment strategy participants are 
employing. We then combine these behavioural patterns with data on demographics, 
personality, social preferences, political ideology, religiosity, and self-reported strategy usage.

![Visual summary of the six economic games.](images/games.png){
  #fig-games
  width="80%"
  apa-note="In the games, Player 2 either steals £0.20 from Player 1 (the focal player) or does nothing. Player 1 is then given the option to punish by paying a certain amount of money to remove money from Player 2 (this money is destroyed). The six games are variants on this general setup, creating situations where (A) Player 2 is still worse off after stealing, (B) Player 2 creates equality by stealing, (C) the computer 'decides' whether Player 2 steals, (D) the fee-fine ratio is 1:1, (E) Player 2 is better off after stealing, and (F) Player 2 steals instead from a third-party."
  }

::: {.landscape}

```{r}
#| label: tbl-strategies
#| tbl-cap: Summary of the different functions for punishment and the behavioural strategies they predict.
#| apa-note: Games A-F are the games employed in the current study (see Methods for more details). In each of the six games, participants are given the opportunity to punish players who 'steal' and those who do not, meaning that participants make twelve punishment decisions in total. Each behavioural strategy implies a unique pattern of punishment across all decisions (see Methods for detailed explanation of strategies). Ticks reflect decisions to punish, crosses reflect decisions to not punish. In column headers, payoffs at the first stage (above) and the second stage (below) are denoted as P1-P2 (or P2-P3 [P1] for Game F) where participants take the role of P1 and P2 is the target of punishment. AI = advantageous inequity, DI = disadvantageous inequity.

tar_read(tableStrategies) %>%
  flextable() %>%
  set_header_labels(
    Function = "Function",
    `Behavioural strategy` = "Behavioural strategy",
    a = "Steal \n 50-30",
    b = "No steal \n 70-10",
    c = "Steal \n 50-50",
    d = "No steal \n 70-30",
    e = "Steal \n 50-50",
    f = "No steal \n 70-30",
    g = "Steal \n 50-50",
    h = "No steal \n 70-30",
    i = "Steal \n 50-70",
    j = "No steal \n 70-50",
    k = "Steal \n 50-90 [100]",
    l = "No steal \n 70-70 [100]"
  ) %>%
  add_header_row(
    values = c(
      "",
      "Game A \n (AI) \n 70-10",
      "Game B \n (Equal) \n 70-30",
      "Game C \n (Computer) \n 70-30",
      "Game D \n (1:1 Fee Fine) \n 70-30",
      "Game E \n (DI) \n 70-50",
      "Game F \n (Third Party) \n 70-70 [100]"
    ),
    colwidths = rep(2, 7)
  ) %>%
  theme_apa() %>%
  width(width = c(1, 2, rep(0.6, 12))) %>%
  fontsize(size = 9, part = "all") %>%
  line_spacing(part = "all") %>%
  align(i = NULL, j = 1:2, align = "left", part = "all")
```

:::

@@

The overall pattern of punitive behaviour in the six economic games was very 
similar across both countries (@fig-decisions). Participants were generally more 
likely to punish targets who stole compared to targets who did not steal 
(multilevel logistic regression; _b_ = 
`r papaja::printnum(mlm1["otherTook","Estimate"])`, standard error = 
`r papaja::printnum(mlm1["otherTook","Std. Error"])`, _p_ 
`r papaja::printp(mlm1["otherTook","Pr(>|z|)"])`). Participants were also more 
likely to punish when targets' stealing behaviour generated inequalities, 
specifically in Games E and F (_b_ = 
`r papaja::printnum(mlm2["DI","Estimate"])`, SE = 
`r papaja::printnum(mlm2["DI","Std. Error"])`, 
_p_ `r papaja::printp(mlm2["DI","Pr(>|z|)"])`).

```{r, fig.height=5, fig.width=8}
#| label: fig-decisions
#| fig-cap: Overall pattern of punitive behaviour across all six economic games, split by country.
#| fig-align: center
#| apa-note: AI = advantageous inequity, DI = disadvantageous inequity.

tar_read(plotDecisions)
```

<br>

We classified participants into a particular strategy if their behaviour across 
all twelve decisions matched our behavioural predictions shown in 
@tbl-strategies exactly. @tbl-strategycounts shows the proportion of 
participants following each strategy, with N/A used to represent participants 
who did not fit exactly into any particular strategy type. Overall, 
`r papaja::printnum(mean(tar_read(dExc)$strategy != "N/A")*100, digits = 0)`%
of our participants could be classified exactly into one of the strategies. The
most common strategy in both countries was to never punish across any of the
games. The next most common strategies were those that care about
minimising payoff differences (avoid disadvantageous inequity, egalitarian). 
Less common were the behaviour shaping strategies (deterrent, norm-enforcing),
the retributive strategy, and the competitive strategies (seek advantageous 
inequity, competitive). Although participants often punished targets who did not 
steal in the six games (@fig-decisions), no participants followed the antisocial
strategy by exclusively punishing targets who did not steal across _all_ games.

```{r}
#| label: tbl-strategycounts
#| tbl-cap: Counts and proportions of participants following each punishment strategy exactly, split by country.
#| apa-note: N/A implies that participants were unable to be classified exactly into any of the punishment strategies. AI = advantageous inequity, DI = disadvantageous inequity.

tar_read(tableStrategyCounts) %>%
  flextable() %>%
  set_header_labels(
    Strategy = "Strategy",
    a = "N",
    b = "Prop",
    c = "N",
    d = "Prop"
  ) %>%
  add_header_row(
    values = c(
      "",
      "United Kingdom\n(N = 1014)",
      "United States\n(N = 996)"
    ),
    colwidths = c(1, 2, 2)
  ) %>%
  autofit() %>%
  theme_apa() %>%
  line_spacing(part = "all") %>%
  align(i = NULL, j = 1, align = "left", part = "all")
```

<br>

To further investigate the strategies that participants were following, we 
examined the most common patterns of punitive behaviour across all twelve 
decisions. @supptbl-patterns shows the proportion of participants following the
25 most common behavioural patterns, including, where appropriate, the 
predetermined strategies from @tbl-strategies. In both countries, a common 
pattern of behaviour not captured by any of the strategies was punishing only 
when the target stole in the third-party game (Game F). Punishment in this game
is consistent with an egalitarian motive, as stealing produces unequal outcomes,
but third-party punishment here is also consistent with norm-enforcing and 
competitive motives (see @tbl-strategies). Other common behavioural patterns not 
captured by our strategies included punishing whenever the target stole across 
all games and always punishing in every game irrespective of the targets' 
behaviour.

@@

To deal with this complexity and include all observed data in our frequency 
estimates, we fitted a Bayesian latent state model to the data. This model 
assumes that the nine strategies in @tbl-strategies (plus a "random choice" 
strategy that chooses randomly for each decision) are the only latent strategies 
and that these are instantiated into observed behaviour according to the logic 
in @tbl-strategies with some probability of implementation error (i.e., an 
intention to punish is implemented as non-punishment and vice versa). Averaging 
over all strategies and incorporating the possibility of implementation errors,
the model estimates the probability of participants following any particular 
strategy, given the observed data.

The posterior estimates from the model are presented in @fig-model1b. The 
posterior probabilities for each strategy did not differ between the two 
countries. In both countries, the never punish strategy had the highest 
probability, followed by the egalitarian strategy. The norm-enforcing and seek 
advantageous inequity strategies were the next most likely, with higher 
posterior estimates than the competitive and antisocial strategies. None of the 
other strategies differed in their posterior estimates. The same general pattern 
emerged when we analysed the full dataset without pre-registered exclusions 
(@suppfig-model1a).
